---
title: "Characterizing Quartzite Points along Ume river valley, Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
institute:
  - UmU: Umeå University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/phd-files/COURSES/DIALPAST/Quantitative_research/MyTestProject/analysis/templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/phd-files/COURSES/DIALPAST/Quantitative_research/MyTestProject/analysis/templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Here is an abstract
keywords: |
  XRF; quartzite; pca
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Settlement and mobility is very complex.


# Background

Forsberg (I don't have Zotero) performed a PCA analysis on the artefact assemblages collected from inland hunter-gatherer sites along North Swedish river systems in the mid 80s. Based on the analysis performed on the morphometrics Forsberg grouped and classified the sites as either residential or activity sites. The resulting settlement pattern was used to test hypotheses of group size....

# Methods

XRF measurements were sampled on quartzite points stored at the collections of Västerbotten and Skellefteå museums. Samples were taken on the surface of the artefacts using a Niton XL 5 portable XRF. 

Data was cleaned... (promise).

Then I did this...


# Results

<!-- Here's some example analysis code: -->

```{r get-data}
library(tidyverse)
library(rrtools)
library(here)
library(infer)
library(FactoMineR)
library(factoextra)

# import raw data
points <- read.csv2(here("analysis/data/raw_data/xrf.csv"), sep = ";", dec = ".")

# Function for replacing NA with mean of variable
f1 <- function(vec) {
  m <- mean(vec, na.rm = TRUE)
  vec[is.na(vec)] <- m
  return(vec)
}

# Use cbind to bind untreated columns and then apply mean function on XRF data 
points_mean = cbind(points[1:27], apply(points[,c(28:58)],2,f1))

# Removes columns that are all NA
points_no_na <- points_mean[, unlist(lapply(points_mean, function(x) !all(is.na(x))))]

# Now that we have found our columns without variance let's get rid of them
# However the result of the apply is not numeric so we have to cast the variable
points_prep <- points_no_na[ - as.numeric(which(apply(points_no_na, 2, var) == 0))]
```

The below figure \@ref(fig:screeplot) visualises the explained variance of the 5 first principal components (PC). As can be seen most of the variance is explained in the first two PCs.  
```{r screeplot, fig.cap="Screeplot for the five first principal components"}
#Prep data for PCA, select only the compositional data
points_pca <- points_prep %>% 
  select(reading_no,Mg:Ba) %>%  
  column_to_rownames(var = "reading_no")

#Compute PCA
res.pca <- PCA(points_pca, graph = FALSE)

#Draw screeplot of PCs
fviz_screeplot(res.pca, 
               ncp = 5,
               title = "First five principal components")
```


Taking a closer look at the first two components, figures \@ref(fig:scree-pc1) and  \@ref(fig:scree-pc2) shows the 10 main contributing variables to each component.  
```{r scree-pc1, fig.cap="Contributions of variables to PC 1"}
#Screeplot of contributing variables to PC1
fviz_contrib(res.pca,
             choice = "var",
             fill = "black",
             color = "black",
             axes = 1,
             top = 10,
             title = "Contribution of variables to PC 1") +
  theme_bw()
```
  
  
    
```{r scree-pc2, fig.cap="Contributions of variables to PC 2"}
#Screeplot of contributing variables to PC2
fviz_contrib(res.pca,
             choice = "var",
             fill = "black",
             color = "black",
             axes = 2,
             top = 10,
             title = "Contribution of variables to PC 1") +
  theme_bw()
```
  
The result of the PCA analysis can be seen in figure \@ref(fig:pca-biplot). Most of the measurements group in a larger cluster, demonstrating little difference in their elemental composition in regards to the  color of the quartzite.  

```{r pca-biplot, fig.height=8,fig.width=10, fig.cap="Biplot of PCA analysis with PC1 and PC2, no clear clustering can be seen related to the color of the material"}
#Draw biplot to visualize result
fviz_pca_biplot(res.pca, axes = c(1, 2),
                geom.ind = "point", # show points only (but not "text")
                pointshape = 16, 
                geom.var = c("point", "text"),
                col.var = "black",
                label = "var",
                palette = "jco",
                col.ind = points_prep$hue, #color by sample hue
                repel = TRUE,
                obs.scale = 1, var.scale = 1,
                legend.title = "Hue",
                title = "PCA - Quartzite points",
                theme_bw(base_size = 40))
```

# Discussion

As quartzite is a metamorphic rock formed from quartz-rich sandstone it is not surprising to see that Si features a high contribution to PC 1. 

# Conclusion


# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
